---
title: "Group Assignment 1 - Time series decomposition"
format: html
editor: source
---

**Released:** Session 7, Monday 28 September 2026.
**Due:** Sunday 25 October 2026, 23:59, on Blackboard.

Four weeks. The deadline falls after the midterm on purpose: revising for the exam and building a
decomposition from scratch are different kinds of work, and they should not compete for the same week.

This is group work. It receives detailed written feedback, unlike the weekly homework.

```{r}
library(fpp3)
```

## The data

`aus_retail` holds monthly retail turnover for every Australian state and industry. The code below
takes one of those series: liquor retailing in Victoria, from January 1990 on.

```{r}
liquor <-
  aus_retail |>
  # 1. One state, one industry
  filter(State == "Victoria", Industry == "Liquor retailing") |>
  # 2. Keep the period from 1990 onward
  filter(year(Month) >= 1990) |>
  # 3. Drop the key columns, leaving the index and the series
  select(Month, Turnover)

autoplot(liquor, Turnover) +
  labs(y = "Turnover ($ million)",
       title = "Liquor retailing turnover, Victoria")
```

Read the plot before you write any code. How the size of the seasonal swing changes as the series
grows is the thing that decides Part 3, so form an expectation now and check it at the end.

# Part 1

Decompose the series with classical decomposition, using the `classical_decomposition()` function
from the fpp3 library. Do it twice:

- an additive decomposition, stored in a dataframe called `dcmp_additive`
- a multiplicative decomposition, stored in a dataframe called `dcmp_multiplicative`

```{r}
dcmp_additive <-
  liquor |>
  # YOUR CODE GOES HERE

dcmp_additive |> autoplot()

dcmp_multiplicative <-
  liquor |>
  # YOUR CODE GOES HERE

dcmp_multiplicative |> autoplot()
```

# Part 2

Now produce the same two decompositions by writing the algorithm yourself, following the four steps
in the notebook `07_D_ClassicalDecomposition_fromScratch`. Build one dataframe per scheme, and keep
every intermediate column in it.

- additive: `manual_dcmp_add`
- multiplicative: `manual_dcmp_mult`

Every component you compute must match the Part 1 result **exactly**. Check as you go rather than at
the end: build a step, compare it against the library, and only then build the next step on top of
it. `all.equal()` is the function to use for these checks.

## Step 1 - Trend estimation

Use the appropriate moving average to create a column called `trend` in each dataframe, matching the
trend that `classical_decomposition()` produces. The choice of moving average depends on the seasonal
period, and the notebook explains which one applies here.

```{r}
manual_dcmp_add <-
  liquor |>
  # YOUR CODE GOES HERE


manual_dcmp_mult <-
  liquor |>
  # YOUR CODE GOES HERE
```

## Step 2 - Detrended series

Remove the trend from the series. The additive and multiplicative schemes do this differently.

```{r}
manual_dcmp_add$detrended <-  # YOUR CODE GOES HERE

manual_dcmp_mult$detrended <- # YOUR CODE GOES HERE
```

## Step 3 - The seasonal component

### Step 3.1 - Unadjusted seasonal component

Build a separate dataframe holding the seasonal pattern of each decomposition, one value per month
of the year.

**Name things exactly as follows:**

- additive: a dataframe called `df_seasonal_add`, with the unadjusted seasonal component in a column
  called `S_unadj`
- multiplicative: a dataframe called `df_seasonal_mult`, with the unadjusted seasonal component in a
  column called `S_unadj`

```{r}
df_seasonal_add <-
  manual_dcmp_add |>
  # YOUR CODE GOES HERE

df_seasonal_mult <-
  manual_dcmp_mult |>
  # YOUR CODE GOES HERE
```

### Step 3.2 - Adjusted seasonal component

Adjust `S_unadj` in each case, and store the result in a new column called `S`. The additive scheme
and the multiplicative scheme are adjusted to different targets. Make sure you know which target
applies to which, and why.

```{r}
df_seasonal_add <-
  df_seasonal_add |>
  # YOUR CODE GOES HERE

df_seasonal_mult <-
  df_seasonal_mult |>
  # YOUR CODE GOES HERE
```

### Step 3.3 - Repeat the seasonal pattern across the whole series

Use `left_join()` to carry the `S` column of `df_seasonal_add` into `manual_dcmp_add`, and the `S`
column of `df_seasonal_mult` into `manual_dcmp_mult`.

```{r}
manual_dcmp_add <-
  manual_dcmp_add |>
  # YOUR CODE GOES HERE

manual_dcmp_mult <-
  manual_dcmp_mult |>
  # YOUR CODE GOES HERE
```

## Step 4 - The remainder

Create a column called `random` in `manual_dcmp_add` and in `manual_dcmp_mult`.

```{r}
manual_dcmp_add$random <- # YOUR CODE GOES HERE

manual_dcmp_mult$random <- # YOUR CODE GOES HERE
```

## Final check: rebuild the original series from your components

Do this for both schemes. Two things to do in each case:

1. Reconstruct the original series from your components.
2. Compare the reconstruction against the original series.

One complication. The moving average produces no values at the two ends of the series, so `trend` and
`random` both start and end with NAs. Compare only over the stretch where your decomposition has no
NAs:

- Build a logical mask that is `TRUE` wherever your decomposition has a value.
- Apply that same mask to the original series and to your reconstruction, so both cover the same
  stretch.

The code below is a reminder of how a mask like that works on a plain vector.

```{r}
# REMINDER: HOW TO MASK NAs OUT OF A VECTOR IN R.

# toy vector for the demonstration
v <- c(NA, NA, 10, 12, 15, 18, 20, 25, 30, 35, NA, NA)

# build the mask
mask <- !is.na(v)

# the mask itself
mask

# the elements that survive it
v[mask]
```


```{r}
# YOUR CODE GOES HERE
```

## Draw the decomposition

Plot your manual decomposition with ggplot commands. Do not use `autoplot()`. Pick either the
additive or the multiplicative case.

Aim for what the fpp3 library prints when you call `autoplot()` on a decomposition, apart from the
grey scale bars on the left of each panel. Those bars are worth understanding, and they come up in
class.

```{r}
# YOUR CODE GOES HERE
```

# Part 3 - Which decomposition is better?

Session 7 gives two criteria for a good decomposition, both of them about the remainder. Apply both
to your two decompositions and decide which scheme suits this series.

Answer all three of the following:

1. **Criterion 1, as numbers.** Report the variance of the trend, the seasonal component and the
   remainder, for both schemes. Then state the one reason these six numbers cannot be compared
   directly across the two schemes, and give a quantity built from them that can be.
2. **Criterion 2, as a picture and a count.** Plot the correlogram of the remainder for both schemes,
   out to lag 24. Count the spikes that cross the blue bounds in each plot and report the two counts.
   Counting off the plot is all that is asked for here.
3. **The verdict.** Name the better scheme. Support it with one number from each criterion, and tie
   it to a feature you can point at in the time plot of the original series.

```{r}
# YOUR CODE GOES HERE
```

------

**YOUR ANSWER GOES HERE. MAX 150 WORDS.**

------
