---
title: "Group Assignment 1 - Time Series Decomposition"
format: html
editor: source
---

**Released:** Session 7, Monday 28 September 2026.
**Due:** Sunday 25 October 2026, 23:59, on Blackboard.

Four weeks, and the deadline falls after the midterm on purpose: revising for the exam and
building a decomposition from scratch are different activities, and they should not compete
for the same week.

This is group work and it receives detailed written feedback, unlike the weekly homework.

```{r}
library(fpp3)
```

The code below creates the a10 dataframe. It details the expenditure on anti diabetic drugs in the Australian healthcare system.

```{r}
a10 <- 
  # 1. Dataframe containing monthly medicare expenses per type of drug
  PBS %>%
  
  # 2. Filter for antidiabetic drigs
  filter(ATC2 == "A10") %>%
  
  # 3. select subset of columns
  select(Month, Concession, Type, Cost) %>%
  
  # 4. Add the total expenditure per month
  index_by(Month) %>% 
  summarise(TotalC = sum(Cost)) %>%
  
  # 5. Scale to millions
  mutate(Cost = TotalC / 1e6)

# 6. Plot
autoplot(a10, Cost) +
  labs(y = "$ (millions)",
       title = "Australian antidiabetic drug sales")
```

# Part 1

Decompose the time series using classical decomposition. Do it using the fpp3 library function `classical_decomposition()`. Perform the decomposition twice: 

- First use an additive approach and store the result in a dataframe called `dcmp_additive`. 
- Then a multiplicative approach and store the result in a dataframe called `dcmp_multiplicative`

```{r}
dcmp_additive <- 
  a10 %>% 
  # YOUR CODE GOES HERE

dcmp_additive %>% autoplot()

dcmp_multiplicative <- 
  a10 %>% 
  # YOUR CODE GOES HERE

dcmp_multiplicative %>% autoplot()
```

# Part 2

Obtain the same decompositions programming the algorithm using yourself using the description in notebook 04_D. Create a separate dataframe for each decomposition variant (additive and multuplicative). Store all the columns containing the intermediate results in these dataframes.

- For the additive case, name the dataframe `manual_dcmp_add`
- For the multiplicative case, name the dataframe `manual_dcmp_mult`

At any intermediate point you consider appropriate, you may perform as many sanity checks as you consider appropriate by comparing with the decompositions provided by the `fpp3` library you computed before. All the components of your manual decompositions must match EXACTLY with the components of the decompositions obtained in part 1. To perform these checks, I recommend using the function `all.equal()`.

## Step 1 - Trend estimation

Use appropriate moving averages to create a new column within both dataframes called `trend` that matches exactly the trend produced by `classical_decomposition()` function of the fpp3 library. Use the algorithm described in notebook 04_D to attain this.

You may use the function `all.equal()` at the end to compare your result to the results of Part 1

```{r}
manual_dcmp_add <- 
  a10 %>% 
  # YOUR CODE GOES HERE


manual_dcmp_mult <- 
  a10 %>% 
  # YOUR CODE GOES HERE  
```

## Step 2 - Detrended series

Obtain the detrended series using the result of step 1

```{r}
manual_dcmp_add$detrended <-  # YOUR CODE GOES HERE

manual_dcmp_mult$detrended <- # YOUR CODE GOES HERE
```

## Step 3 - Compute seasonal component

#### Step 3.1 - Unadjusted seasonal component

Create a new dataframe containing the pattern of the unadjusted seasonal component of each decomposition.

**Present the results exactly as follows**: 
- For the additive decomposition, create a dataframe called `df_seasonal_add`. Name the column of the unadjusted seasonal component `S_unadj`.
- For the multiplicative decomposition, create a dataframe called `df_seasonal_mult`. Name the column of the unadjusted seasonal component `S_unadj`.

```{r}
df_seasonal_add <- 
  manual_dcmp_add %>% 
  # YOUR CODE GOES HERE

df_seasonal_mult <-
  manual_dcmp_mult %>% 
  # YOUR CODE GOES HERE
```

#### Step 3.2 - Adjusted seasonal components

Perform the necessary operations to adjust `S_unadj` for both the additive and the multiplicative decompositions. Store the result in a new column `S` within `df_seasonal_add` (`df_seasonal_mult`) for the additive (multiplicative) case.

```{r}
df_seasonal_add <- 
  df_seasonal_add %>% 
  # YOUR CODE GOES HERE

df_seasonal_mult <- 
  df_seasonal_mult %>% 
  # YOUR CODE GOES HERE
```

#### Step 3.2 - Chain the adjusted seasonal components to cover the entire timespan of the time series

To do this, use `left_join()` to bring the pattern of the S column of `df_seasonal_add` (`df_seasonal_mult`) into `manual_dcmp_add` (`manual_dcmp_mult`)

```{r}
manual_dcmp_add <- 
  manual_dcmp_add %>% 
  # YOUR CODE GOES HERE

manual_dcmp_mult <- 
  manual_dcmp_mult %>% 
  # YOUR CODE GOES HERE
```

## Step 4 - Obtain the random component

Create a new column called `random` within `manual_dcmp_add` and `manual_dcmp_mult`

```{r}
manual_dcmp_add$random <- # YOUR CODE GOES HERE

manual_dcmp_mult$random <- # YOUR CODE GOES HERE
```

## Final check: check that the original time series can be reconstructed exactly from the components you created

For clarity, you need to:

1. Reconstruct the original time series from your components.
2. Compare your reconstructed series with the original time series.

There is a caveat: 

- Because the moving averages did not generate values at the extremes of the series, the trend and the random of your decompositions will contain NAs at the beginning and the end. 
- You should make the comparison between your decompositions and your time series only at the points where your decompositions have no NAs. 
- Use a logical mask to select only the region of values from the original time series where your decomposition has no NAs. Use that same logical mask on your decomposition as well, to exclude NAs.
- As a reminder, given a vector `v` in R, the code below illustrates how to produce a logical mask that allows you to select only values without NAs

```{r}
# REMINDER: HOW TO CREATE A MASK TO FILTER NAs FROM A VECTOR IN R.

# toy vector for demonstration
v <- c(NA, NA, 10, 12, 15, 18, 20, 25, 30, 35, NA, NA)

# Produce logical mask
mask <- !is.na(v)

# Print the mask
mask 

# Select only the elements of the vector that do not have NAs
v[mask]
```


```{r}
# YOUR CODE GOES HERE
```

## Graphical depiction

Depict your manual decomposition using gg_plot commands. Do not resort to `autoplot()`. Pick either the additive or the multiplicative case.

The output should be what is printed by the `fpp3` library when using `autoplot()` on a decomposition, except for the vertical bars (professor will clarify in class, so pay attention).

```{r}
# YOUR CODE GOES HERE
```

# Part 3 - Judging which decomposition is better

Compute a quantitative metric to compare both decompositions. Use it to assess which decomposition is better. Draw a conclusion.

```{r}
# YOUR CODE GOES HERE
```