---
title: "session6_hw_tsa"
format: html
---


```{r}
library(fpp3)
```


Exercise 1

Using us_retail_employment, compute the 2×12 moving average of Employed.
The data is monthly, so the seasonal period is m = 12, which is even. Work out .before and.after for each of the two imbalanced windows before you type anything.
Plot the result over the original series.

```{r}
# Filter the series and select relevant columns
us_retail_employment <- 
  us_employment |>
  filter(year(Month) >= 1990, Title == "Retail Trade") |>
  select(-Series_ID)


us_retail_ma <- us_retail_employment |>
  mutate(
    MA_12_R = slider::slide_dbl(Employed, mean,
                                .before = 5, .after = 6, .complete = TRUE),
    MA_12_L = slider::slide_dbl(Employed, mean,
                                .before = 6, .after = 5, .complete = TRUE),
    MA_2x12 = 0.5 * (MA_12_R + MA_12_L)
  )




```

```{r}
us_retail_ma |>
  select(Month, Employed, MA_12_R, MA_12_L, MA_2x12) |>
  head(10)
```
```{r}
us_retail_ma |>
  select(Month, Employed, MA_12_R, MA_12_L, MA_2x12) |>
  tail(10)
```

At the beginning:
- MA_12_R needs 5 months before + current + 6 after, so it becomes available after 5 missing rows.
- MA_12_L needs 6 months before + current + 5 after, so it becomes available after 6 missing rows.
- Since MA_2x12 needs both, it starts only when both exist.
At the end:
- One of them runs out 5 months before the end.
- The other runs out 6 months before the end.
- So the combined MA_2x12 also has missing values near the end.

So the final MA_2x12 loses first 6 and last 6 rows. 

```{r}
us_retail_ma |>
  autoplot(Employed, colour = "grey") +
  geom_line(aes(y = MA_2x12), colour = "brown") +
  labs(
    title = "US retail employment with 2x12 moving average",
    y = "People employed"
  )
```


